One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-20 paper/appendix.tex
first draft done, waiting on timeres results

In fact, we can simplify Eq. \ref{tbg_3d_2} in the general case by
using $\delta\rho = - \rho\nabla\cdot\bm{\xi} -
\bm{\xi}\cdot\nabla\rho$, giving
\begin{align}\label{tbg_general}
  T_\mathrm{BG} = \frac{m}{2}\frac{p}{\rho
    c_s^2}\imag\left[\rho\left(\nabla\cdot\bm{\xi}\right)\bm{\xi}^*\cdot\nabla
  c_s^2\right].  
\end{align} 
For a barotropic fluid $p=p(\rho)$, the function $c_s^2$ can be
taken as constant (Eq. \ref{iso_cond}) for which $T_\mathrm{BG}$
vanishes. When there is a forced temperature gradient,
Eq. \ref{tbg_general} indicates there is angular momentum exchange
between compressible perturbations ($\nabla\cdot\bm{\xi}\neq0$) and
the background disc if there is motion along the temperature
gradient ($\bm{\xi}\cdot\nabla c_s^2 \neq 0$).

>>>>

In fact, we can simplify Eq. \ref{tbg_3d_2} in the general case by
using $\delta\rho = - \rho\nabla\cdot\bm{\xi} -
\bm{\xi}\cdot\nabla\rho$, giving
\begin{align}\label{tbg_general}
  T_\mathrm{BG} = \frac{m}{2}\frac{p}{\rho
    c_s^2}\imag\left[\rho\left(\nabla\cdot\bm{\xi}\right)\bm{\xi}^*\cdot\nabla
  c_s^2\right].  
\end{align} 
For a barotropic fluid $p=p(\rho)$, the function $c_s^2$ can be
taken as constant (Eq. \ref{iso_cond}) for which $T_\mathrm{BG}$
vanishes. When there is a forced temperature gradient,
Eq. \ref{tbg_general} indicates a torque is applied to compressible
perturbations ($\nabla\cdot\bm{\xi}\neq0$) if there is motion along
the temperature gradient ($\bm{\xi}\cdot\nabla c_s^2 \neq 0$).

### 2014-11-20 paper/discussion.tex
first draft done, waiting on timeres results

\subsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-armed spiral  
to act as an external potential for the wider disc. This is
analogous to disc-satellite interaction 
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}. Here, we estimate the magnitude of this effect using
basic results from disc-planet theory \citep[see, e.g.][and references 
therein]{papaloizou07}.

@@

%and here we are interested in the
%potential it produces at $R>\overline{R}$  
%This potential is associated with a mass
%$M_\mathrm{ring}$ contained in   is the mass in $R\in[R_1,R_2]$.

@@

If we associate the external potential with angular momentum 
$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can obtain a
rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right)^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\
  &\times \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L + 2\left[1 -
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2.
\end{align}
%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \overline{R}\right)^2}. 
%\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 5\times10^{-4}\Omega_p. 
\end{align}
%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}
%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-armed spiral.  
For the co-rotation torque, we use the result
\begin{align}
  \Gamma_c = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}   
\end{align}
for a Keplerian disc, where subscript $c$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right)
\left[b_{1/2}^1(\beta_c)\right]^2   
\end{align}
%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
 %     - \overline{R}\right)^2}.   
%\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
\end{align}

@@

% both are positive -> destabilizing in principle

%\begin{align}
%  \gamma_C\sim 0.010\Omega_p,
%\end{align}
%which is again small compared to the measured growth rates.

%not important for our models, but may be important if outer disc is
%self-gravitating too

%expressions for linear response, but have massive "planet" -> but CR
%and LR far from where spiral is

%required disc structure (for confined SG spiral)

%\subsection{Speculations}
%speculations - long term balance. forced temp gradient drives
%instability (disc gains ang mom, spiral loses ang mom). but if spiral
%shocks and dissiple, it dumps negative angular momentum (disc gains
%negative ang mom). balance between gain from forced temp gradient and
%loss due to dissiplatoin

%other source of m=1 disturbance -> eccentric modes (adams, PJ06,
%hopkins, tremaine)

>>>>

\subsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-armed spiral  
to act as an external potential for the wider disc. This is
analogous to disc-satellite interaction 
\citep{goldreich79}. Here, we estimate the magnitude of this effect
using basic results from disc-planet theory \citep[see, e.g.][and
references therein]{papaloizou07}.

@@

@@

If we associate the external potential with angular momentum 
$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can obtain a
rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right) 
  ^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\ 
  &\times
  \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L
    + 2\left[1 - 
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2. 
\end{align}
%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \overline{R}\right)^2}. 
%\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 5\times10^{-4}\Omega_p. 
\end{align}
%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}
%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-armed spiral.  
For the co-rotation torque, we use the result
\begin{align}
  \Gamma_c = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}     
\end{align}
for a Keplerian disc, where subscript $c$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right) 
\left[b_{1/2}^1(\beta_c)\right]^2    
\end{align}
%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
 %     - \overline{R}\right)^2}.   
%\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
\end{align}

@@

The torque exerted on the disc at the OLR by an
external potential is positive, while that at co-rotation 
depends on the gradient of potential vorticity there
\citep{goldreich79}. For our disc models with surface density
$\Sigma\propto R^{-2}$ in the outer disc, this co-rotation torque is
positive. Interpreting the one-armed spiral as an external potential
for the outer disc, this means that the spiral loses
angular momentum by launching density waves with positive angular momentum at
the OLR, and by applying a positive co-rotation torque on the disc. In
principle, this interaction is destabilising because the one-armed
spiral has negative angular momentum \citep{lin11b}. As estimated
above, this effect is small in our simulations, but could be
important for other parameter regimes.

%also transport ang mom outwards

### 2014-11-20 paper/results.tex
first draft done, waiting on timeres results

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-armed spiral
we measure a co-rotation radius and growth rate  
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo.ps}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo.ps}
  \caption{Normalised radial wavenumber of the $m=1$ spiral in 
    Fig. \ref{2d_fargo_viz}.\label{fargo_wavenumber}} 
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo.ps} 
  \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
    Fig. \ref{2d_fargo_viz}. For a given real mode frequency, the
    dispersion relation for local density waves, Eq. \ref{dispersion},
    permits purely wave-like solutions in regions where $\nu^2 - 1 +
    Q^{-2}>0$.    
    \label{fargo_qbarrier}} 
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo.ps} 
  \caption{Rate of change of the $m=1$ wave angular momentum as defined by
    Eq. \ref{baroclinic_torque_int} (solid) compared to the torque
    exerted on the wave associated with the background temperature
    gradient (dotted). 
    \label{fargo_angmom_ex}} 
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq.ps}   
  \caption{Evolution of the $m=1$ spiral amplitude as a function of
    the imposed sound-speed gradient $q$. The maximum value of the
    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
    \label{fargo_varq}} 
\end{figure}

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates
This confirms that the temperature gradient is responsible for the
development of the one-armed spirals observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}
%maybe extend the low q simulations to get better rates?

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varh.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    disc aspect-ratio $h$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varh_growth}} 
\end{figure}

>>>>

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-armed spiral
we measure a co-rotation radius and growth rate  
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo}
  \caption{Normalised radial wavenumber of the $m=1$ spiral in 
    Fig. \ref{2d_fargo_viz}.\label{fargo_wavenumber}} 
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo} 
  \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
    Fig. \ref{2d_fargo_viz}. For a given real mode frequency, the
    dispersion relation for local density waves, Eq. \ref{dispersion},
    permits purely wave-like solutions in regions where $\nu^2 - 1 +
    Q^{-2}>0$.    
    \label{fargo_qbarrier}} 
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo} 
  \caption{Rate of change of the $m=1$ wave angular momentum as defined by
    Eq. \ref{baroclinic_torque_int} (solid) compared to the torque
    exerted on the wave associated with the background temperature
    gradient (dotted). 
    \label{fargo_angmom_ex}} 
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq}   
  \caption{Evolution of the $m=1$ spiral amplitude as a function of
    the imposed sound-speed gradient $q$. The maximum value of the
    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
    \label{fargo_varq}} 
\end{figure}

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates
This confirms that the temperature gradient is responsible for the
development of the one-armed spirals observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}
%maybe extend the low q simulations to get better rates?

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varh}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    disc aspect-ratio $h$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varh_growth}} 
\end{figure}

### 2014-11-20 paper/results_3d.tex
first draft done, waiting on timeres results

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ampmax3d.ps}
  \caption{Evolution of the maximum $m=1$ density component in  $r\in[R_1,R_2]$
    in the 3D simulations. Results from discs with a 
    temperature gradient ($q=1$) and a strictly isothermal disc
    ($q=0$) are shown.  
    \label{3d_ampmax}}   
\end{figure}

@@

\begin{figure}
  \includegraphics[scale=0.47,clip=true,trim=0cm 0.79cm 0cm
  0cm]{figures/pdisk_rz_023_sg.ps}  
  \includegraphics[scale=0.47,clip=true,trim=0cm 0cm 0cm
  0.64cm]{figures/pdisk_rz_023_nsg.ps}  
  \caption{The $m=1$ density component in the meridional plane in the 
    PLUTO simulation. The slices are taken at the azimuth of   
    $\mathrm{max}[\Delta\rho_1(r,\pi/2,\phi)]$. Arrows represent the vector 
    $(v_r/R_0,-v_\theta/rh\sin^2{\theta})$. The top (bottom) panel corresponds
    to the inner (outer) portions of the disc. 
    \label{3d_rz}} 
\end{figure}

>>>>

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ampmax3d}
  \caption{Evolution of the maximum $m=1$ density component in  $r\in[R_1,R_2]$
    in the 3D simulations. Results from discs with a 
    temperature gradient ($q=1$) and a strictly isothermal disc
    ($q=0$) are shown.  
    \label{3d_ampmax}}   
\end{figure}

@@

\begin{figure}
  \includegraphics[scale=0.47,clip=true,trim=0cm 0.79cm 0cm
  0cm]{figures/pdisk_rz_023_sg}  
  \includegraphics[scale=0.47,clip=true,trim=0cm 0cm 0cm
  0.64cm]{figures/pdisk_rz_023_nsg}  
  \caption{The $m=1$ density component in the meridional plane in the 
    PLUTO simulation. The slices are taken at the azimuth of   
    $\mathrm{max}[\Delta\rho_1(r,\pi/2,\phi)]$. Arrows represent the vector 
    $(v_r/R_0,-v_\theta/rh\sin^2{\theta})$. The top (bottom) panel corresponds
    to the inner (outer) portions of the disc. 
    \label{3d_rz}} 
\end{figure}

### 2014-11-20 paper/summary.tex
first draft done, waiting on timeres results

\section{Summary and conclusions}\label{summary}
In this paper, we have carried out direct numerical hydrodynamic
simulations of radially structured, self-gravitating locally
isothermal discs.     
We find such systems can be unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which lead to the development of one-armed 
trailing spirals that persist for at least $O(10^2)$ orbits. The 
spiral pattern speed is smaller than the local disc rotation, making
it almost stationary. The growth rates are $O(10^{-2}\Omega)$ which
gives a characteristic growth time of $O(10P_0)$. 
%While this is slow
%compared to the local rotation, it is dynamical at co-rotation.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to verify the growth of  
one-armed spirals in our disc models is due to the imposed temperature
profile and disc structure. The spiral is confined between two
$Q$-barriers owing to a surface density bump. Growth rates increased
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations.  
We find no significant vertical 
motion in 3D simulations, which indicates the instability is
two-dimensional.

We applied angular momentum conservation within linear theory to
explain our numerical results. A  
forced temperature gradient introduces a torque on linear 
perturbations. We called this the background torque because it
represents an exchange of angular momentum between the background disc
and the perturbations. In the local approximation, we showed that this
background torque is negative for trailing spirals, which enforces  
low-frequency $m=1$ modes because they are associated with negative
angular momentum. We therefore interpret the one-armed spiral instability
observed in our simulations as a neutral, tightly-wound $m=1$ mode
being destabilised by the imposed temperature gradient.

%works in 2D/3D
%forced gradient feedback 
%neutral mode destabilized      
\subsection{Speculations and future work}
%accretion??
%balancing heating and cooling 
In the deeply non-linear regime, the one-armed spiral may 
shock and dissipate, depositing negative angular momentum onto
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
forced, it may be possible to achieve a balance between the gain of negative
angular momentum through the forced temperature gradient, and the gain
of positive angular momentum through shock dissipation. In order to
study this possibility, improved numerical models are needed to ensure 
total angular momentum conservation in long-term simulations.

%numerical simulations to include cooling/heating
A locally isothermal disc represents the limit of infinitely short
cooling (and heating) timescales, so the disc temperature instantly
returns to its initial value. %so we don't go to deeply non-linear
                             %regime 
This assumption can be relaxed by including an energy
equation with thermal relaxation. Preliminary FARGO simulations
indicate a relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is needed for
the one-armed spiral to develop. This will be presented in a follow-up
study. %prelim sims show they are long lasting

%non-sg sims, Be stars, eccentric discs
% We stress that destabilisation through the temperature gradient
% does not explicitly depend on self-gravity. In principle, this effect
% will destabilise low-frequency $m=1$ trailing spirals regardless of
% its origin.
In our models and interpretation, self-gravity coupled with the disc 
structure permits a neutral one-armed mode, which is then
destabilised. Other origins of a neutral
one-armed spiral should be investigated. One possibility is in Be star discs
\citep{rivinius13}, for which 
one-armed oscillations may explain long-timescale variations in their
emission lines \citep[see e.g.][and references
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.

>>>>

\section{Summary and conclusions}\label{summary}
In this paper, we have carried out direct numerical hydrodynamic
simulations of radially structured, self-gravitating locally
isothermal discs.     
We find such systems can be unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which lead to the development of one-armed 
trailing spirals that persist for at least $O(10^2)$ orbits. The 
spiral pattern speed is smaller than the local disc rotation, and
growth rates are $O(10^{-2}\Omega)$ which gives a characteristic
growth time of $O(10)$ orbits.   
%While this is slow
%compared to the local rotation, it is dynamical at co-rotation.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to verify the growth of  
one-armed spirals in our disc models is due to the imposed temperature
profile and disc structure. The spiral is confined between two
$Q$-barriers owing to a surface density bump. Growth rates increased
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations.  
We find the instability behaves similarly in 2D and 3D. However, in 3D
the spiral disturbance can become radially global away from 
the midplane.

We applied angular momentum conservation within linear theory to
explain our numerical results. We find a   
forced temperature gradient introduces a torque on linear 
perturbations. We called this the background torque because it
represents an exchange of angular momentum between the background disc
and the perturbations. This offers a previously unexplored pathway to
instability.

In the local approximation, we showed that this background torque is
negative for trailing spirals in discs with a fixed temperature or
sound-speed profile that decrease outwards. This torque enforces  
low-frequency $m=1$ modes because they are associated with negative
angular momentum. We therefore interpret the one-armed spiral instability
observed in our simulations as an initially neutral, tightly-wound
$m=1$ mode being destabilised by the imposed temperature gradient.

\subsection{Speculations and future work}
%numerical simulations to include cooling/heating
A locally isothermal equation of state represents the limit of
infinitely short cooling (and heating) timescales, so the disc
temperature instantly returns to its initial value when perturbed.  
This assumption can be relaxed by including an energy equation with
thermal relaxation. Preliminary FARGO simulations
indicate a thermal relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is
needed for the one-armed spiral to develop. However, this value may
depend on disc parameters, and will be explored in a follow-up  
study.

% prelim sims show they are long lasting 
% balancing heating and cooling 
% should not get fragmentation (CR away from spiral) 
In the deeply non-linear regime, the one-armed spiral may 
shock and deposit negative angular momentum onto
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
maintained, it may be possible to achieve a balance between the gain
of negative angular momentum through the temperature gradient, and the
gain of positive angular momentum through shock dissipation. We remark  
that fragmentation is unlikely because the co-rotation radius 
is outside the bulk of the spiral arm, in the non-self-gravitating
portion of the disc \citep{durisen08}. In order
to study these possibilities, improved numerical models are needed to
ensure total angular momentum conservation on timescales much longer
than that considered in this paper.

% non-sg sims, Be stars, eccentric discs
% We stress that destabilisation through the temperature gradient
% does not explicitly depend on self-gravity. In principle, this effect
% will destabilise low-frequency $m=1$ trailing spirals regardless of
% its origin.

In our models and interpretation, self-gravity coupled with the disc 
structure permits a neutral one-armed mode, which is then
destabilised by the temperature gradient. Other origins of the neutral
one-armed spiral should be investigated. One possibility is in Be star discs
\citep{rivinius13}, for which 
one-armed oscillations may explain long-timescale variations in their
emission lines \citep[see e.g.][and references
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.
